Two subtle problems with over-representation analysis

### 2024-06-14 manuscript/manuscript.Rmd
reproducibility with master script and dockerfile

library("DiagrammeR")
library("RColorBrewer")
library("kableExtra")
#library("gt")

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
To demonstrate the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected, as suggested by a previous report [@Tarca2013-xh].
These gene lists were subjected to ORA using the `enricher` function of clusterProfiler
v4.10.0, using a minimum set size of 5 and a no maximum set size.

To systematically assess the impact of the background problem, we conducted ORA with
`enricher` using the different gene set libraries described above for the seven RNA-seq
datasets.
`enricher` was used with the same parameters as above, and we devised a simple workaround
to the background problem, which is to append the entire background list as a gene set to the
library.
This forces `enricher` to retain all background genes.
The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
Jaccard index was used to compare original and corrected analyses.

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset based on p-value.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running `enricher`.
After `enricher` using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
enricher, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

@@

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets (FDR<0.05) over seven independent datasets detected with
clusterProfiler's `enricher` function with and without correcting the background problem.
Separate tests for up- and down-regulated gene lists were conducted.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler's `enricher` function, which suffers from
the two aforementioned problems and `fora` which does not.
In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
prototypical FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.2, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.69
with added variance of 0.5.
In comparison, the precision of fora was relatively stable around 0.97 until added variance
reached 0.5, where it dropped to 0.88.
The recall of ORA methods dropped suddenly when added variance was 0.2 or greater.
fgsea showed stable precision of 0.93 or greater when added variance was 0.5 or lower, which
dropped slightly to 0.91 with added variance of 0.6.
fgsea recall was better than either ORA tool across all levels of added variance.
This result shows that these two subtle problems result in poorer precision for
clusterProfiler, which becomes more severe with higher levels of noise.
Overall, fgsea had similar mean precision to fora (0.94), but the higher recall of fgsea
meant it had a higher overall accuracy measured by F1 score (0.83) as compared to fora and
clusterProfiler (0.61 and 0.64 respectively).

![**Figure 3. Comparison of clusterProfiler, fora and fgsea accuracy with simulated data.**
Precision, recall and F1 score are shown with the addition of different amounts of random
noise to expression levels.](../analysis/sim/fig3_sim.png "Figure 3 shows that fora is more
precise than clusterProfiler, and fgsea has higher recall than either ORA methods.")

@@

Although the purpose of this work was to describe two subtle problems with ORA, this work
highlights some major drawbacks.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Secondly ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis)[@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.

>>>>

library("kableExtra")

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
To demonstrate the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected, as suggested by a previous report [@Tarca2013-xh].
These gene lists were subjected to ORA using clusterProfiler's `enricher` function
v4.10.0, using a minimum set size of 5 and a no maximum set size.

To systematically assess the impact of the background problem, we conducted ORA with
clusterProfiler using the different gene set libraries described above for the seven RNA-seq
datasets.
ClusterProfiler was used with the same parameters as above, and we devised a simple workaround
to the background problem, which is to append the entire background list as a gene set to the
library.
This forces clusterProfiler to retain all background genes.
The ORA analyses were then filtered for significant sets using a FDR threshold of 0.05.
Jaccard index was used to compare original and corrected analyses.

@@

For this test, 2000 of the top up and down regulated genes were selected from each
differential expression dataset based on p-value.
A workaround for the FDR problem was devised by filtering the gene sets by the presence of
at least two genes in the background list prior to running clusterProfiler.
After clusterProfiler using the parameters above, the number of gene sets in the results was
quantified, and compared to the number that should have been reported using the two gene
limit.
The difference between these numbers (*n*) represents the number of missing gene sets.
To account for these tests, *n* values of 1 were appended to the p-values obtained by
clusterProfiler, followed by FDR correction in R, to obtain the properly corrected FDR values.
As above, analysis of seven datasets was done with nine gene set libraries, the significant
sets were selected and the Jaccard statistics were collected for each run to compare original
and corrected analysis.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0.

@@

![**Figure 1A. Effect of correcting the background problem.** Figures represent the mean number
of statistically significant sets (FDR<0.05) over seven independent datasets detected with
clusterProfiler with and without correcting the background problem.
Separate tests for up- and down-regulated gene lists were conducted.
The percent increase in significant sets are shown.](../analysis/fig1_ORAsig.png "Figure 1A
shows that correcting the background bug improves the number of significant sets between 12 and
229%.")

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler default and with mitigations for
problems described above.
In addition, these data underwent ORA with fora which does not suffer from these two
problems.
For comparison fgsea was used to compare the performance of these two ORA methods against a
prototypical FCS method.
These methods were tested with different levels of introduced variation and the mean
precision, recall and F1 scores were recorded (**Figure 3A**).
As anticipated, mitigating the background problem improved recall (by 1.6%), however it caused
a similar-sized decrease in precision.
Addressing the FDR problem improved precison by 9.6%, but the recall dropped by 8.5%.
Addressing both problems improved precision by 8.2% but decreased recall by 5.6%.
Under these conditions, the FDR problem affected results more severely than the background
problem.
Results obtained for clusterProfiler after implementing the two mitigations were identical to
fora to three significant figures.
fgsea recorded precision of 0.937, similar to the ORA methods, but recall was far superior,
with fgsea scoring 22.4% higher than default clusterProfiler and 28% higher than fora.
As a result, fgsea's overall accuracy as defined by F1 index was 21.2% higher than fora and
18.4% higher than default clusterProfiler.

When looking at the accuracy profile at different levels of added noise (**Figure 3B**), we
see that ORA techniques have superior precision when the added noise is low (<0.2), but the
precision of ORA techniques drops at higher levels of added noise.
The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
The precision of fgsea was relatively stable despite high levels of added noise, ranging from
0.95 to 0.91.
The recall of fgsea was superior to all ORA approaches across all levels of added noise.

![**Figure 3A. Impact of correcting these two problems on accuracy of simulated expression
data.**
Precision, recall and F1 scores are the mean of trials at a range of added noise.
CP; clusterProfiler, CP BG fix; clusterProfiler with a workaround to the background problem,
CP FDR fix; clusterProfiler with a workaround to the FDR problem,
CD BG and FDR fix; clusterProfiler with mitigations for both background and FDR problems,
fora; ORA function of the fgsea package,
fgsea; an FCS method similar to GSEA.
](../analysis/sim/fig3_bars.png "Figure 3A shows that fora is more
precise than clusterProfiler, and fgsea has higher recall than ORA methods.")

![**Figure 3B. Accuracy of clusterProfiler, fora and fgsea with simulated data with various
levels of added noise.**
Precision, recall and F1 score are shown with the addition of different amounts of random
noise to expression levels.
The purple line for clusterProfiler with mitigations for both problems is not visible as the
performance was identical to fora.
](../analysis/sim/fig3_sim.png "Figure 3B shows that ORA methods have deteriorating accuracy
with noisy data as compared to fgsea, a FCS method.")

@@

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.

Secondly ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis)[@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.
